Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2019-04-29 paper/paper.tex
cite File S4, S5 in text as requested by JVI

\documentclass[lineno]{asm-article}

@@

We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

@@

The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
Visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription using incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
\FIG{mutations}A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.

@@

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{transcriptomics}B.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{frac_has_gene}). 
To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{ISG}, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).

@@

\clearpage
\section{Supplemental Material}

>>>>

\documentclass[]{asm-article}

@@

We next used PacBio sequencing (\FIG{workflow}D, \SUPPFILE{primer_sequences}) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

@@

The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes} (see also \SUPPFILE{genotypes}).
Visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription using incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
\FIG{mutations}A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement} (see also \SUPPFILE{mutations}).
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.

@@

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{transcriptomics}B.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{frac_has_gene}). 
To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{ISG}, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).
The raw data are in \SUPPFILE{genotypes} and \SUPPFILE{mutations}.

@@

\newpage
\setcounter{page}{1}
\section{Supplemental Material}

### 2019-04-29 paper/paper_final_version_JVI_no-figs-supp.tex
version for JVI final submission
cite File S4, S5 in text as requested by JVI

\input{asm-biblatex.tex}
\bibliography{references}

@@

We next used PacBio sequencing (Fig. 2D) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (Fig. S9).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (Fig. S10), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

@@

The 150 cells for which we called the full viral genotypes are shown in Fig. 4.
Visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in Fig. 4, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in Fig. 4 into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
Fig. 5A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.

@@

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at https://github.com/jbloomlab/IFNsorted\_flu\_single\_cell/blob/master/monocle\_analysis.ipynb, and is also provided as File S8.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in Fig. 3B.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (Fig. 3C).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (Fig. 3D and Fig. S4). 
To identify cells that were co-infected with both viral barcodes (Fig. 3F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in Fig. 3G and Fig. S6, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in Fig. 4 summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (https://github.com/wilkox/gggenes).

@@

>>>>

\input{asm-biblatex.tex}

\bibliography{references}

@@

We next used PacBio sequencing (Fig. 2D, File S3) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (Fig. S9).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (Fig. S10), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

@@

The 150 cells for which we called the full viral genotypes are shown in Fig. 4 (see also File S4).
Visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in Fig. 4, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in Fig. 4 into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
Fig. 5A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement} (see also File S5).
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.

@@

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at https://github.com/jbloomlab/IFNsorted\_flu\_single\_cell/blob/master/monocle\_analysis.ipynb, and is also provided as File S8.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in Fig. 3B.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (Fig. 3C).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (Fig. 3D and Fig. S4). 
To identify cells that were co-infected with both viral barcodes (Fig. 3F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in Fig. 3G and Fig. S6, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in Fig. 4 summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (https://github.com/wilkox/gggenes).
The raw data are in File S4 and File S5.

@@

\printbibliography
